---
title: "Modeling"
author: "Team VM"
format: 
    html:
        toc: true
        toc-depth: 3
        toc-location: left
        toc-title: "Table of Contents"
        embed-resources: true
---

```{r}
# Libraries
library(tidyverse)
library(dplyr)
library(tibble)
library(caret)
library(pROC)
library(themis)
library(ipred)
library(randomForest)
library(gbm)
# remotes::install_github("souravc83/fastAdaboost")
library(klaR)
library(fastAdaboost)
library(kernlab)
library(glmnet)
library(doParallel)
library(foreach)
library(ggplot2)

set.seed(42)
RANDOM_SEED <- 42
```

```{r}
# Read file
df <- read.csv("Data_Files/Train.csv") %>%
  dplyr::select(-ID)

df_holdout = read.csv("Data_Files/Test.csv")
holdout_ids <- df_holdout$ID
```

## Data Preparation

### Log Transform Skewed Variables
```{r}
# log1p is used to handle zero values
df$Prior_Season_Grand_Prix_Wins_Last5GP <-
  log1p(df$Prior_Season_Grand_Prix_Wins_Last5GP)

df_holdout$Prior_Season_Grand_Prix_Wins_Last5GP <-
  log1p(df_holdout$Prior_Season_Grand_Prix_Wins_Last5GP)
```

### Stratified Train-Test Split

```{r}
# Split into train and test datasets
set.seed(RANDOM_SEED)

train_index <- createDataPartition(
  df$Constructor_Champion,
  p = 0.7,
  list = FALSE,
  times = 1
)

train_df <- df[train_index, ]
test_df  <- df[-train_index, ]


cat("Training set size:", nrow(train_df), "\n")
cat("Test set size:", nrow(test_df), "\n")
```

```{r}
# Verify stratified splitting
print(mean(train_df$Constructor_Champion))
print(mean(test_df$Constructor_Champion))
```

### Convert Target Variable to Factor

```{r}
# Binary predictor variables
binary_vars <- c(
  "New_Name",
  "New_Team",
  "UK_Based",
  "Major_Reg_Change",
  "Defending_Driver_Champ",
  "Defending_Constructor_Champ",
  "Returning_Driver_Lineup"
)

train_df[binary_vars] <- lapply(train_df[binary_vars], factor)
test_df[binary_vars]  <- lapply(test_df[binary_vars], factor)
df[binary_vars] <- lapply(df[binary_vars], factor)
df_holdout[binary_vars] <- lapply(df_holdout[binary_vars], factor)

# Target variable
train_df$Constructor_Champion <- factor(
  train_df$Constructor_Champion,
  levels = c(1, 0),
  labels = c("Yes", "No")
)

test_df$Constructor_Champion <- factor(
  test_df$Constructor_Champion,
  levels = c(1, 0),
  labels = c("Yes", "No")
)

df$Constructor_Champion <- factor(
  df$Constructor_Champion,
  levels = c(1, 0),
  labels = c("Yes", "No")
)
```

## Base Model Training

### Parallel Processing Setup

```{r}
# Use one fewer core than available
n_cores <- max(1, parallel::detectCores() - 1)

cl <- makeCluster(n_cores)
registerDoParallel(cl)

cat("Using", foreach::getDoParWorkers(), "cores\n")
```

### Cross-Validation Setup

```{r}
set.seed(RANDOM_SEED)

# without SMOTE
cv_control <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

# with SMOTE
cv_control_smote <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  sampling = "smote",
  savePredictions = "final",
  allowParallel = TRUE
)
```

### Logistic Regression Models

```{r}
set.seed(RANDOM_SEED)

log_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control
)
```

```{r}
set.seed(RANDOM_SEED)

log_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control_smote
)
```

### Decision Tree Models

```{r}
set.seed(RANDOM_SEED)

tree_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "rpart",
  metric = "ROC",
  trControl = cv_control_smote
)
```

```{r}
set.seed(RANDOM_SEED)

tree_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "rpart",
  metric = "ROC",
  trControl = cv_control
)
```

### Ensemble Models

#### Bagging

```{r}
set.seed(RANDOM_SEED)

bag_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "treebag",
  metric = "ROC",
  trControl = cv_control
)
```

```{r}
set.seed(RANDOM_SEED)

bag_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "treebag",
  metric = "ROC",
  trControl = cv_control_smote
)
```

#### Random Forest

```{r}
set.seed(RANDOM_SEED)

rf_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "rf",
  metric = "ROC",
  ntree = 500,
  importance = TRUE,
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

rf_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "rf",
  metric = "ROC",
  ntree = 500,
  importance = TRUE,
  trControl = cv_control_smote,
  tuneLength = 3
)
```

#### Gradient Boosting

```{r}
set.seed(RANDOM_SEED)

gbm_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "gbm",
  metric = "ROC",
  verbose = FALSE,
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

gbm_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "gbm",
  metric = "ROC",
  verbose = FALSE,
  trControl = cv_control_smote,
  tuneLength = 3
)
```

#### AdaBoost

```{r}
set.seed(RANDOM_SEED)

ada_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "AdaBoost.M1",
  metric = "ROC",
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

ada_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "AdaBoost.M1",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### Regularization Models

#### Ridge Logistic Regression

```{r}
set.seed(RANDOM_SEED)

ridge_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control,
  tuneGrid = expand.grid(
    alpha = 0,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

```{r}
set.seed(RANDOM_SEED)

ridge_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneGrid = expand.grid(
    alpha = 0,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

#### Lasso Logistic Regression

```{r}
set.seed(RANDOM_SEED)

lasso_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control,
  tuneGrid = expand.grid(
    alpha = 1,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

```{r}
set.seed(RANDOM_SEED)

lasso_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneGrid = expand.grid(
    alpha = 1,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

#### Elastic Net Logistic Regression

```{r}
set.seed(RANDOM_SEED)

enet_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

enet_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### Naive Bayes

```{r}
set.seed(RANDOM_SEED)

nb_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "nb",
  metric = "ROC",
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

nb_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "nb",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### K-Nearest Neighbors

```{r}
set.seed(RANDOM_SEED)

knn_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "knn",
  metric = "ROC",
  preProcess = c("center", "scale"),
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

knn_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "knn",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### Support Vector Models

#### SVM Linear

```{r}
set.seed(RANDOM_SEED)

svm_linear_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "svmLinear",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

svm_linear_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "svmLinear",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3
)
```

#### SVM Radial

```{r}
set.seed(RANDOM_SEED)

svm_radial_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "svmRadial",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

svm_radial_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "svmRadial",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### Neural Network

```{r}
set.seed(RANDOM_SEED)

nnet_model <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "nnet",
  metric = "ROC",
  preProcess = c("center", "scale"),
  trControl = cv_control,
  tuneLength = 3,
  trace = FALSE,
  maxit = 1000
)
```

```{r}
set.seed(RANDOM_SEED)

nnet_model_smote <- train(
  Constructor_Champion ~ .,
  data = train_df,
  method = "nnet",
  metric = "ROC",
  preProcess = c("center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3,
  trace = FALSE,
  maxit = 1000
)
```

### Base Model Comparison

```{r}
get_class_metrics <- function(model, model_name, test_data, target_col = "Constructor_Champion") {
  
  preds <- predict(model, test_data)
  probs <- predict(model, test_data, type = "prob")[, "Yes"]
  
  cm <- confusionMatrix(
    preds,
    test_data[[target_col]],
    positive = "Yes"
  )
  
  model_auc <- auc(
    roc(
      response = test_data[[target_col]],
      predictor = probs,
      levels = c("No", "Yes"),
      direction = "<"
    )
  )
  
  data.frame(
    Model = model_name,
    AUC = round(as.numeric(model_auc), 6),
    Accuracy = round(cm$overall["Accuracy"], 6),
    Precision = round(cm$byClass["Precision"], 6),
    Recall = round(cm$byClass["Recall"], 6),
    F1 = round(cm$byClass["F1"], 6),
    Specificity = round(cm$byClass["Specificity"], 6)
  )
}
```

```{r}
model_comparison <- bind_rows(
  get_class_metrics(log_model, "Logistic Regression", test_df),
  get_class_metrics(log_model_smote, "Logistic Regression + SMOTE", test_df),
  get_class_metrics(tree_model, "Decision Tree", test_df),
  get_class_metrics(tree_model_smote, "Decision Tree + SMOTE", test_df),
  get_class_metrics(bag_model, "Bagging", test_df),
  get_class_metrics(bag_model_smote, "Bagging + SMOTE", test_df),
  get_class_metrics(rf_model, "Random Forest", test_df),
  get_class_metrics(rf_model_smote, "Random Forest + SMOTE", test_df),
  get_class_metrics(gbm_model, "Gradient Boosting", test_df),
  get_class_metrics(gbm_model_smote, "Gradient Boosting + SMOTE", test_df),
  get_class_metrics(ada_model, "AdaBoost", test_df),
  get_class_metrics(ada_model_smote, "AdaBoost + SMOTE", test_df),
  get_class_metrics(ridge_model, "Ridge Logistic Regression", test_df),
  get_class_metrics(ridge_model_smote, "Ridge Logistic Regression + SMOTE", test_df),
  get_class_metrics(lasso_model, "Lasso Logistic Regression", test_df),
  get_class_metrics(lasso_model_smote, "Lasso Logistic Regression + SMOTE", test_df),
  get_class_metrics(enet_model, "Elastic Net", test_df),
  get_class_metrics(enet_model_smote, "Elastic Net + SMOTE", test_df),
  get_class_metrics(nb_model, "Naive Bayes", test_df),
  get_class_metrics(nb_model_smote, "Naive Bayes + SMOTE", test_df),
  get_class_metrics(knn_model, "KNN", test_df),
  get_class_metrics(knn_model_smote, "KNN + SMOTE", test_df),
  get_class_metrics(svm_linear_model, "SVM Linear", test_df),
  get_class_metrics(svm_linear_model_smote, "SVM Linear + SMOTE", test_df),
  get_class_metrics(svm_radial_model, "SVM Radial", test_df),
  get_class_metrics(svm_radial_model_smote, "SVM Radial + SMOTE", test_df),
  get_class_metrics(nnet_model, "Neural Network", test_df),
  get_class_metrics(nnet_model_smote, "Neural Network + SMOTE", test_df)
) %>%
  arrange(desc(AUC), desc(Precision))

knitr::kable(
  model_comparison,
  digits = 4,
  caption = "Model Comparison Results"
)
```

### End Processing Setup

```{r}
stopCluster(cl)
registerDoSEQ()

cat("Parallel processing stopped\n")
```

## Final Model

```{r}
set.seed(RANDOM_SEED)

svm_radial_grid <- expand.grid(
  sigma = c(0.0001, 0.0005, 0.001, 0.005, 0.01, 0.05, 0.1),
  C = c(0.25, 0.5, 1, 2, 4, 8, 16)
)
```

```{r}
set.seed(RANDOM_SEED)

svm_radial_tuned <- train(
  Constructor_Champion ~ .,
  data = df,
  method = "svmRadial",
  metric = "ROC",
  trControl = cv_control,
  preProcess = c("zv", "center", "scale"),
  tuneGrid = svm_radial_grid
)
```

```{r}
svm_radial_tuned$bestTune
```

### Generate Final Predictions

```{r}
holdout_probs <- predict(
  svm_radial_tuned,
  newdata = df_holdout,
  type = "prob"
)

holdout_preds <- predict(
  svm_radial_tuned,
  newdata = df_holdout
)

y2026_predictions <- data.frame(
  ID = holdout_ids,
  Probability_Yes = holdout_probs$Yes
) %>%
  arrange(desc(Probability_Yes)) %>%
  mutate(
    Predicted_Standing = row_number()
  )

write.csv(
  y2026_predictions,
  "Data_Files/y2026_predictions.csv",
  row.names = FALSE
)
```