---
title: "Modeling"
author: "Team VM"
format: 
    html:
        toc: true
        toc-depth: 3
        toc-location: left
        toc-title: "Table of Contents"
        embed-resources: true
---

```{r}
# Libraries
library(caret)
library(pROC)
library(themis)
library(ipred)
library(randomForest)
library(gbm)
library(klaR)
library(adabag)
library(kernlab)
library(glmnet)
library(doParallel)
library(foreach)
library(tidyverse)
library(DT)

RANDOM_SEED <- 42
```

```{r}
# Read file
df <- read.csv("Data_Files/Train.csv")
df_holdout <- read.csv("Data_Files/Test.csv")
```

## Data Preparation

### Log Transform Skewed Variables
```{r}
# log1p is used to handle zero values
df$Prior_Season_Grand_Prix_Wins_Last5GP <-
  log1p(df$Prior_Season_Grand_Prix_Wins_Last5GP)

df_holdout$Prior_Season_Grand_Prix_Wins_Last5GP <-
  log1p(df_holdout$Prior_Season_Grand_Prix_Wins_Last5GP)
```

### Expanding Window Cross-Validation

```{r}
# Seasons in chronological order
seasons <- sort(unique(df$Season))

# Store training and validation indices
train_index <- list()
test_index  <- list()

# Start with the first 15 seasons (2000-2014)
for (i in 16:length(seasons)) {

  train_seasons <- seasons[1:(i - 1)]
  test_season   <- seasons[i]

  train_index[[paste0("Fold", i - 15)]] <-
    which(df$Season %in% train_seasons)

  test_index[[paste0("Fold", i - 15)]] <-
    which(df$Season == test_season)
}
```

```{r}
# verify fold creation
length(train_index)
length(test_index)

sapply(train_index, length)
sapply(test_index, length)
```

```{r}
fold_check <- tibble(
  Fold = names(test_index),
  Train_Start = sapply(train_index, function(x) min(df$Season[x])),
  Train_End = sapply(train_index, function(x) max(df$Season[x])),
  Validation_Season = sapply(test_index, function(x) unique(df$Season[x])),
  Training_Rows = sapply(train_index, length),
  Validation_Rows = sapply(test_index, length)
)

#knitr::kable(fold_check)
datatable(fold_check)
```

### Convert Binary Variables to Factors

```{r}
# Binary predictor variables
binary_vars <- c(
  "New_Name",
  "New_Team",
  "UK_Based",
  "Major_Reg_Change",
  "Defending_Driver_Champ",
  "Defending_Constructor_Champ",
  "Returning_Driver_Lineup"
)

df[binary_vars] <- lapply(df[binary_vars], factor)
df_holdout[binary_vars] <- lapply(df_holdout[binary_vars], factor)

# Target variable
df$Constructor_Champion <- factor(
  df$Constructor_Champion,
  levels = c(1, 0),
  labels = c("Yes", "No")
)

# Ensure holdout factor levels match training
for (v in binary_vars) {
  df_holdout[[v]] <- factor(
    df_holdout[[v]],
    levels = levels(df[[v]])
  )
}
```

### Create Modeling Datasets

```{r}
# Create modeling datasets
df_model <- df %>%
  dplyr::select(-ID, -Season)

df_holdout_model <- df_holdout %>%
  dplyr::select(-ID, -Season, -Constructor_Champion)

holdout_ids <- df_holdout$ID
```

## Base Model Training

### Parallel Processing Setup

```{r}
# Use one fewer core than available
n_cores <- max(1, parallel::detectCores() - 1)

cl <- makeCluster(n_cores)
registerDoParallel(cl)

cat("Using", foreach::getDoParWorkers(), "cores\n")
```

### Cross-Validation Setup

```{r}
set.seed(RANDOM_SEED)

# without SMOTE
cv_control <- trainControl(
  method = "cv",
  index = train_index,
  indexOut = test_index,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE,
  verboseIter = TRUE
)
```

```{r}
set.seed(RANDOM_SEED)

# with SMOTE
cv_control_smote <- trainControl(
  method = "cv",
  index = train_index,
  indexOut = test_index,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  sampling = "smote",
  savePredictions = "final",
  allowParallel = TRUE,
  verboseIter = TRUE
)
```

### Logistic Regression Models

```{r}
set.seed(RANDOM_SEED)

log_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  preProcess = "zv",
  trControl = cv_control
)
```

```{r}
set.seed(RANDOM_SEED)

log_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  preProcess = "zv",
  trControl = cv_control_smote
)
```

### Decision Tree Models

```{r}
set.seed(RANDOM_SEED)

tree_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "rpart",
  metric = "ROC",
  trControl = cv_control_smote
)
```

```{r}
set.seed(RANDOM_SEED)

tree_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "rpart",
  metric = "ROC",
  trControl = cv_control
)
```

### Ensemble Models

#### Bagging

```{r}
set.seed(RANDOM_SEED)

bag_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "treebag",
  metric = "ROC",
  trControl = cv_control
)
```

```{r}
set.seed(RANDOM_SEED)

bag_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "treebag",
  metric = "ROC",
  trControl = cv_control_smote
)
```

#### Random Forest

```{r}
set.seed(RANDOM_SEED)

rf_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "rf",
  metric = "ROC",
  ntree = 500,
  importance = TRUE,
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

rf_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "rf",
  metric = "ROC",
  ntree = 500,
  importance = TRUE,
  trControl = cv_control_smote,
  tuneLength = 3
)
```

#### Gradient Boosting

```{r}
set.seed(RANDOM_SEED)

gbm_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "gbm",
  metric = "ROC",
  verbose = FALSE,
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

gbm_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "gbm",
  metric = "ROC",
  verbose = FALSE,
  trControl = cv_control_smote,
  tuneLength = 3
)
```

#### AdaBoost

```{r}
set.seed(RANDOM_SEED)

ada_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "AdaBoost.M1",
  metric = "ROC",
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

ada_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "AdaBoost.M1",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### Regularization Models

#### Ridge Logistic Regression

```{r}
set.seed(RANDOM_SEED)

ridge_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control,
  tuneGrid = expand.grid(
    alpha = 0,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

```{r}
set.seed(RANDOM_SEED)

ridge_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneGrid = expand.grid(
    alpha = 0,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

#### Lasso Logistic Regression

```{r}
set.seed(RANDOM_SEED)

lasso_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control,
  tuneGrid = expand.grid(
    alpha = 1,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

```{r}
set.seed(RANDOM_SEED)

lasso_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneGrid = expand.grid(
    alpha = 1,
    lambda = 10^seq(-3, 1, length = 3)
  )
)
```

#### Elastic Net Logistic Regression

```{r}
set.seed(RANDOM_SEED)

enet_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control,
  tuneGrid = expand.grid(
    alpha = c(0.25, 0.50, 0.75),
    lambda = c(0.001, 0.01, 0.1)
  )
)
```

```{r}
set.seed(RANDOM_SEED)

enet_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control_smote,
  tuneGrid = expand.grid(
    alpha = c(0.25, 0.50, 0.75),
    lambda = c(0.001, 0.01, 0.1)
  )
)
```

### Naive Bayes

```{r}
set.seed(RANDOM_SEED)

nb_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "nb",
  metric = "ROC",
  preProcess = "zv",
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

nb_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "nb",
  metric = "ROC",
  preProcess = "zv",
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### K-Nearest Neighbors

```{r}
set.seed(RANDOM_SEED)

knn_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "knn",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

knn_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "knn",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### Support Vector Models

#### SVM Linear

```{r}
set.seed(RANDOM_SEED)

svm_linear_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "svmLinear",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

svm_linear_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "svmLinear",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3
)
```

#### SVM Radial

```{r}
set.seed(RANDOM_SEED)

svm_radial_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "svmRadial",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control,
  tuneLength = 3
)
```

```{r}
set.seed(RANDOM_SEED)

svm_radial_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "svmRadial",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3
)
```

### Neural Network

```{r}
set.seed(RANDOM_SEED)

nnet_model <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "nnet",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control,
  tuneLength = 3,
  trace = FALSE,
  maxit = 1000
)
```

```{r}
set.seed(RANDOM_SEED)

nnet_model_smote <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "nnet",
  metric = "ROC",
  preProcess = c("zv", "center", "scale"),
  trControl = cv_control_smote,
  tuneLength = 3,
  trace = FALSE,
  maxit = 1000
)
```

### Base Model Comparison

```{r}
get_cv_metrics <- function(model, model_name) {

  best <- model$bestTune

  results <- model$results

  # Keep only the row corresponding to the best tuning parameters
  if (!is.null(best)) {
    for (param in names(best)) {
      results <- subset(results, results[[param]] == best[[param]])
    }
  }

  data.frame(
    Model = model_name,
    AUC = round(results$ROC, 6),
    Sensitivity = round(results$Sens, 6),
    Specificity = round(results$Spec, 6)
  )
}
```

```{r}
model_comparison <- bind_rows(
  get_cv_metrics(log_model, "Logistic Regression"),
  get_cv_metrics(log_model_smote, "Logistic Regression + SMOTE"),
  get_cv_metrics(tree_model, "Decision Tree"),
  get_cv_metrics(tree_model_smote, "Decision Tree + SMOTE"),
  get_cv_metrics(bag_model, "Bagging"),
  get_cv_metrics(bag_model_smote, "Bagging + SMOTE"),
  get_cv_metrics(rf_model, "Random Forest"),
  get_cv_metrics(rf_model_smote, "Random Forest + SMOTE"),
  get_cv_metrics(gbm_model, "Gradient Boosting"),
  get_cv_metrics(gbm_model_smote, "Gradient Boosting + SMOTE"),
  get_cv_metrics(ada_model, "AdaBoost"),
  get_cv_metrics(ada_model_smote, "AdaBoost + SMOTE"),
  get_cv_metrics(ridge_model, "Ridge Logistic Regression"),
  get_cv_metrics(ridge_model_smote, "Ridge Logistic Regression + SMOTE"),
  get_cv_metrics(lasso_model, "Lasso Logistic Regression"),
  get_cv_metrics(lasso_model_smote, "Lasso Logistic Regression + SMOTE"),
  get_cv_metrics(enet_model, "Elastic Net"),
  get_cv_metrics(enet_model_smote, "Elastic Net + SMOTE"),
  get_cv_metrics(nb_model, "Naive Bayes"),
  get_cv_metrics(nb_model_smote, "Naive Bayes + SMOTE"),
  get_cv_metrics(knn_model, "KNN"),
  get_cv_metrics(knn_model_smote, "KNN + SMOTE"),
  get_cv_metrics(svm_linear_model, "SVM Linear"),
  get_cv_metrics(svm_linear_model_smote, "SVM Linear + SMOTE"),
  get_cv_metrics(svm_radial_model, "SVM Radial"),
  get_cv_metrics(svm_radial_model_smote, "SVM Radial + SMOTE"),
  get_cv_metrics(nnet_model, "Neural Network"),
  get_cv_metrics(nnet_model_smote, "Neural Network + SMOTE")
) %>%
  arrange(desc(AUC))

datatable(model_comparison)
```

### End Processing Setup

```{r}
stopCluster(cl)
registerDoSEQ()

cat("Parallel processing stopped\n")
```

## Final Model

### Hyperparameter Tuning

```{r}
set.seed(RANDOM_SEED)

enet_grid <- expand.grid(
  alpha = seq(0, 1, by = 0.1),
  lambda = c(0.0001, 0.001, 0.005, 0.01, 0.05, 0.1, 0.5, 1)
)

enet_tuned <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = cv_control,
  tuneGrid = enet_grid
)

enet_tuned
```

```{r}
plot(enet_tuned)
```

```{r}
enet_tuned$bestTune
```


### Train Final Model

```{r}
set.seed(RANDOM_SEED)

enet_final <- train(
  Constructor_Champion ~ .,
  data = df_model,
  method = "glmnet",
  family = "binomial",
  trControl = trainControl(
    method = "none",
    classProbs = TRUE
  ),
  tuneGrid = enet_tuned$bestTune
)
```

### Generate Final Predictions

```{r}
holdout_probs <- predict(
  enet_final,
  newdata = df_holdout_model,
  type = "prob"
)

holdout_preds <- predict(
  enet_final,
  newdata = df_holdout_model
)

y2026_predictions <- data.frame(
  ID = holdout_ids,
  Predicted_Class = holdout_preds,
  Probability_Yes = holdout_probs$Yes
) %>%
  arrange(desc(Probability_Yes)) %>%
  mutate(
    Predicted_Standing = row_number()
  )

write.csv(
  y2026_predictions,
  "Data_Files/y2026_predictions.csv",
  row.names = FALSE
)

y2026_predictions
```